Predicting detection probabilities for rare and undersampled North American landbirds

### 2025-07-04 article/article_r1.tex
adding revised .tex file and difference file produced by latexdiff
revisions to various parts of the text

		Accurate assessments of population status and trends rely on accounting for probability of detection.
		Testing git-latexdiff.
		However, probability of detection is often unaccounted for or unavailable on a species level. 
		Although methods to estimate detection probability have been developed and implemented for many North American birds, estimates are most accurate for common, easy to detect species. 
		There is therefore a need for methodology that provides detection estimates for species that are rare or undersampled, which generally lack sufficient data for accurate estimates. 
		Here, we take advantage of similarities in detection probabilities among phylogenetically-related species and species with similar traits to estimate detection probabilities in seven species of rare North American birds, using hierarchical Bayesian multi-species removal and distance models. 
		We showed that a multi-species model improved precision of detectability estimates for species with low sample size, and even with species up to 1000 observations.
		However, when no data were available for a species, the multi-species models presented here tended to be overly conservative in predicting detectability. 
		We recommend using these models to 1) provide more precise estimates of detectability for species where there are sufficient data, 2) generate estimates of detectability for species for which we have sparse (but non-zero) data, and 3) improve upon previous estimates of detection probabilities for North American birds by combining this model with survey-level covariates.

@@

\section{Keywords}
\par hierarchical modelling, distance sampling, removal sampling, QPAD, detectability, effective detection radius, cue rate, phylogenetics, bayesian

@@

\par Evidence suggests detectability in birds is at least partially driven by variation in species traits.
\citet{johnston_species_2014} showed that detection distance - and consequently perceptibility - varied with body mass and habitat association in a selection of European landbirds.
In this case, larger body size was associated with longer detection distances, likely because  larger body sizes are easier to detect visually and typically associated with louder and lower-pitched songs \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
Moreover, vegetation structure influences the attenuation of sound, whereby sound diminishes quicker in a closed-forested habitat than more open habitats, thus leading to variation in detectability among species associated with different habitats \citep{waide_tropical_1988, yip_sound_2017}.
\citet{solymos_phylogeny_2018} considered the effects of species traits and phylogeny on both detection distance and cue rate, the latter of which influences the availability of birds for detection.
They found that cue rate is most related to phylogeny, and, to a lesser extent, migratory strategy (i.e., whether a bird was a long-distance migrant or a resident species).
They also found that detection distance was related to species’ traits such as body mass, song pitch, habitat association, and that after accounting for these traits, phylogeny explained very little.

\par \citet{solymos_phylogeny_2018} recommended using these phylogenetic and trait relationships to predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) more broadly.
 By leveraging information provided by species with many detections in available datasets, it may be possible to inform detectability estimates for species with fewer (including potentially zero) detections \citep{sollmann_hierarchical_2016, pacifici_guidelines_2014, zipkin_impacts_2009, white_conservation_2013}.
This could be done in two ways: by making use of shrinkage factors and partial pooling, as done in Bayesian hierarchical models \citep{gelman_what_2021, gelman_bayesian_2006}, and by accounting for phylogenetic relationships and the relative "relatedness" of different species, as done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}.

\par The goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of sample size requirements for estimating detectability, and to derive estimates of detectability for bird species without any data.
To achieve this goal, we aimed to satisfy the following objectives:
\begin{enumerate}
	\item Compare the mean and precision of estimates of cue rate and EDR produced from a multi-species, Bayesian hierarchical QPAD removal model and distance model, respectively, to estimates produced from single-species models that do not share information across species.
	For species with a sufficient sample size, we predicted the mean and precision of cue rate and EDR estimates will be similar from multi- and single-species models \citep{edwards_point_2023}.
	However, for species with relatively small sample sizes, we predict estimates for both cue rate and EDR will be more precise from the multi-species Bayesian model.
	\item Assess the capacity for a multi-species model to produce estimates of cue rate or EDR for species that are data sparse.
	We predicted that a multi-species model will be able to produce better fitting estimates of cue rate or EDR than a single species model, particularly for data sparse species.
	\item Assess the capacity for a multi-species model to make out-of-sample predictions of cue rate or EDR for species which have no data.
	We predicted that a multi-species model will be able to produce estimates of cue rate or EDR that fall within a reasonable range of estimates from related species.
\end{enumerate}

Given these comparisons, we provide a case study of predicting detectability components for seven species for which we had little to no data: Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Spotted Owl (\textit{Strix occidentalis}), Bicknell’s Thrush (\textit{Catharus bicknelli}), LeConte’s Thrasher (\textit{Toxostoma lecontei}), Harris’s Sparrow (\textit{Zonotrichia querula}), Tricolored Blackbird (\textit{Agelaius tricolor}), and Kirtland’s Warbler (\textit{Setophaga kirtlandii}). 	
All seven of these bird species are listed on Partners in Flight’s watchlist, indicating their conservation importance \citep{will_handbook_2020}.

@@

\subsection{Data Collection}
\subsubsection{Point Count Data}
\par We used the same standardized database of point count data that was assembled through the NA-POPS project \cite{edwards_point_2023}. 
This consists of $712,138$ point counts conducted across $292$ projects in Canada and the United States, $422,514$ of which used removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and $522,820$ of which used distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}.
 All point counts were recorded in the database using a standard method developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.

@@

\subsubsection{Species Traits}
\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts for all species considered in this study \citep{billerman_birds_2022}. 
Species were classified as "closed habitat" if their primary habitat was forests, and "open habitat" if their primary habitat was grasslands or shrublands.
Species with specific areas of their range that are only for breeding (i.e., not year-round throughout their range) were classified as "migratory".
We note that for the purposes of this study, altitudinal and latitudinal migrants were classified as migratory, but irruptive species such as crossbills were not.
Body mass for each bird species was obtained from the Elton traits dataset if available \citep{wilman_eltontraits_2014}, and through the Birds of the World account \citep{billerman_birds_2022} if not.

@@

\par For any point count that employed removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and/or distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occurring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
That is, if we have a point count where removal sampling or distance sampling (but not both) was not used, then either $J = 1$ or $K = 1$, respectively; whereas, if we have a point count where both removal sampling and distance sampling was used, then both $J,K \geq 2$.
The case where $J = K = 1$ is where neither removal sampling nor distance sampling was used, and so these data points are not appropriate for the modelling techniques described here.

@@

$$Y_{sij.} \sim multinomial\left(Y_{si..}, \mathbf{\Pi}_{si.}\right).$$

\par For sample $i$, $\mathbf{\Pi}_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $J$ time bands; in essence, it is the mixing parameter.
Let $t_{ij}$ be the maximum time for time band $j$ during sampling event $i$, and let $\phi_s$ be the unknown cue rate for species $s$.
We then have the following calculation of each component $\pi_{sij}$ of the mixing parameter $\mathbf{\Pi}_{si.}$:

\begin{equation*}
	\pi_{sij} = 
	\begin{cases}
		\dfrac{\exp\left\{ -t_{i,j-1}\phi_{s} \right\} - \exp\left\{ -t_{ij}\phi_{s} \right\}}{1 - \exp\left\{ -t_{iJ}\phi_{s} \right\}} & \text{for } j > 1 \\
		1 - \sum_{n = 2}^{J} \pi_{sin} & \text{for } j = 1
	\end{cases}
\end{equation*}

\par We are most interested in estimating $\phi_s$, the cue rate for species $s$. 
To allow for the partial pooling of estimates for each species to be based on a phylogenetic relationship, we can use Gaussian processes, which effectively allow for ``continuous" factors \citep{bernardo_regression_1998, mcelreath_continous_2020}. 
Let $\mathbf{\Phi}$ be the vector of cue rates (of length $s$). 
We then have:

$$\log \mathbf{\Phi} \sim MVN\left( \mathbf{\mu}_{[s]}, \mathbf{\Sigma}_{[s \times s]} \right).$$

\par The covariance function $\mathbf{\Sigma}$ (denoted with subscript $[s \times s]$ to show dimensionality) determines the strength of the partial pooling, and can be any function that contains a measure of distance. 
In this case, we will create a function based on phylogenetic distance $\mathbf{C}_{[s \times s]}$, i.e.,

$$\mathbf{\Sigma}_{[s \times s]} = \mathbf{C}_{[s \times s]}\lambda\sigma.$$

\par Here, $\lambda$ is Pagel's Lambda \citep{pagel_inferring_1999}, which determines the strength of the phylogenetic relationship. 
We set $\lambda = 0.76$, consistent with previous results \cite{solymos_phylogeny_2018}. 
We could have chosen to draw estimates from a distribution of possible $\lambda$ as in \citet{solymos_phylogeny_2018}, but we chose not to for simplicity. 
We note that the suite of species modelled in \citet{solymos_evaluating_2018} was primarily Boreal birds, and so there could be slight differences in the phylogenetic strength of a larger suite of species such as what we model here.
We also estimated the extra variance term and used an exponential prior $\sigma \sim exp(5)$.

\par Finally, the vector of mean cue rate $\mathbf{\mu}_{[s]}$ is a function of the species migratory strategy $\alpha_M$, where the migratory strategy $M$ could either be resident or migrant. 
We set priors on both migratory strategies to be $\alpha_M \sim N(-1, 0.01)$, because on the log scale we would expect cue rate to be negative.

@@

\par For distance modelling, we are only interested in the distance to the cueing bird, and not the time bin that the bird was recorded in, and so we sum counts from each sampling event $i$ over all time bands. 
Thus, we consider counts $Y_{si.k} = \sum_{j=1}^{J}$. 
From \citet{solymos_calibrating_2013}, we model $Y_{si.k}$ as
$$Y_{si.k} \sim multinomial\left(Y_{si..}, \mathbf{\Pi}_{si.}\right).$$

\par For sample $i$, $\mathbf{\Pi}_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $K$ distance bins; in essence, it is the mixing parameter. 
Let $r_{ik}$ be the maximum distance for distance bin $k$ during sampling event $i$, and let $\tau_s$ be the unknown effective detection radius for species $s$. 
We then have the following calculation of each component $\pi_{sik}$ of the mixing parameter $\mathbf{\Pi}_{si.}$:

\begin{equation*}
	\pi_{sik} = 
	\begin{cases}
		\dfrac{f(r_{i,k}, \tau_s) - f(r_{i,k-1}, \tau_s)}{f(r_{i,K}, \tau_s)} & \text{for } k > 1 \\
		1 - \sum_{n = 2}^{K} \pi_{sin} & \text{for } k = 1
	\end{cases}
\end{equation*}

where 
$$f(r,\tau) =  1 - \exp\left\{ -\dfrac{r^2}{\exp\left\{\tau^2\right\}} \right\} .$$

@@

\par In this model, mean effective detection radius for species $s$ (i.e., $\mu_s$) is a function of the following traits as described in \cite{solymos_phylogeny_2018}: an overall intercept term $\alpha_0$; migratory strategy $\alpha_M$, where $M$ can be migratory or resident; habitat preference $\alpha_H$, where $H$ can be open or closed habitat; log body size $\beta_b$; and log song pitch $\beta_p$. 
That is, we have
$$ \mu_s = \alpha_0 + \alpha_{M_s} + \alpha_{H_s} + \beta_b \times \log BodySize_s + \beta_p \times \log SongPitch_s$$
$$\alpha_0 \sim N(0.05, 0.1)$$
$$ \alpha_M, \alpha_H \sim N(0, 0.05)$$
$$ \beta_b \sim N(0.01, 0.005)$$
$$ \beta_p \sim N(-0.01, 0.005) $$
$$\sigma \sim exp(5)$$

@@

\subsection{Model Assessment}
\subsubsection{Objective 1: Single-Species vs. Multi-Species Estimates}
\par To test our prediction that there would be little change between estimates of species with sufficient sizes, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data.
We did not include Woodhouse's Scrub-jay (\textit{Aphelocoma woodhouseii}), Pacific Wren (\textit{Troglodytes pacificus}), or Sagebrush Sparrow (\textit{Artemisiospiza nevadensis}) in the analysis, because these species were not available in the phylogenies from \citet{jetz_global_2012}.
Rufous Hummingbird (\textit{Selasphorus rufus}) was dropped from the distance modelling due to issues with divergent transitions \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
Tricolored Blackbird was only included in the distance model because there was sufficient distance sampling data.
This resulted in a total of 316 species for removal modelling, and 315 species for distance modelling.
We also ran the single species removal and distance models for the same suite of birds.
We then modelled the multi species estimates of cue rate and EDR as functions of the single species estimates of cue rate and EDR, and obtained estimates of slope and y-intercept for this relationship.
If our prediction is true, we would expect to see a slope of 1 and a y-intercept of 0, indicating a 1-to-1 relationship.

@@

\par To test our prediction about improved accuracy, we ran a k-fold cross-validation with 5 folds, stratified by species.
That is, for each of the five cross-validation runs, we held out 20\% of the data for each species for both the multi-species and single-species models. 
We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR, for each cross-validation run, and used these parameter estimates to calculate the estimated log pointwise posterior density (LPPD) of each of the held out data points, which tells us the likelihood of the data point under the estimated parameters.
We then calculated the difference in LPPD for removal and distance models, by subtracting the pointwise LPPD of the single-species model from the LPPD of the multi-species model (i.e., multi-species minus single-species).
Finally, we took the mean and standard error of the difference in LPPD for each species to create a species-specific mean and standard error of difference in LPPD.
With this, positive intervals (i.e., where mean $\pm$ standard error is all positive) would indicate that the multi-species model was preferred, and negative intervals would indicate that the single-species model was preferred, because there would be an overall higher average likelihood of the data points given the parameters in the multi-species model than the single-species model with a positive interval.
If our prediction is true, then we would expect to see majority positive values of the species-specific mean LPPD for species with lower sample sizes.

\subsubsection{Objective 3: Assessment of Out-of-sample Predictions}
\par To test our prediction that a multi-species model would be able to predict cue rates or EDRs for species with no data, we ran a 10-fold leave-species-out cross validation.
That is, for each of the ten cross-validation runs, we held out all data for 10\% of the species for both the multi-species and single-species models.
To determine which species would be left out in each fold of the removal model, we first split the phylogenetic tree into 10 parts using the "cutree" function in base R, which preserves relatedness in the 10 parts of the split tree.
Then, for each of the parts of the tree, we randomly assigned a number between 1 and 10 (drawn from a uniform distribution) to all migratory then all resident species.
This ensured that for each fold, there would be a balance of related and unrelated species left out, as well as a relatively equal balance of migratory and resident species left out.
To determine which species would be left out in each fold of the distance model, we randomly assigned a number between 1 and 10 (drawn from a uniform distribution) to each species within a migratory resident $\times$ habitat preference, which ensured that there would be a good balance of species left out for each group of species.
Because Stan allows draws from a posterior distribution even with no data, we drew values from the posterior distribution of all the held out species for each cross-validation fold, and used mean and standard deviation of those draws as the predicted cue rate or EDR.
We then modelled the predictions of cue rate and EDR obtained with no data as a function of cue rate and EDR obtained with the full data, and obtained estimates of slope and y-intercept for this relationship.
If our prediction is true, then we would expect to have a slope of 1 and a y-intercept of 0, indicating a 1-to-1 relationship.

\subsection{Case Study: Predicting Detection Probabilities for Rare and Undersampled Species}
\par We ran the multi-species removal and distance models again, this time with the minimal data available for seven species of undersampled landbirds, and generated estimates of cue rate and EDR for these seven species.
For Tricolored Blackbird, we had sufficient data points for distance modelling for Objectives 1 - 3, and so we only generated estimates of cue rate here.
For two of the species, Kirtland's Warbler and Bicknell's Thrush, we used the predicted cue rate and EDR to calculate availability and perceptibility, and hence detectability, for a given survey type, and compared these metrics to previous estimates of detectability obtained from other studies.
As a reminder, following derivations in \citet{solymos_calibrating_2013}, detectability is the product of availability and perceptibility.
Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as

@@

\par Out-of-sample predictions of EDR also tended to be underestimated compared to estimates obtained with full data (y-intercept = $4.46$, slope = $0.72$; Figure \ref{fig:species_cv}B).
Of particular note was the fact that the model did not generate any out-of-sample predictions greater than 115 m, nor less than 69 m.
In fact, the model was particularly conservative in generating estimates for each of the migrant x habitat groups, in that there was very little variation within each of the groups; any variation that did occur would have been due to differences in song pitch and body size (Figure \ref{fig:distance_cv_traits}).

>>>>

		Accurate assessments of population status and trends rely on accounting for probability of detection.
		However, probability of detection is often unaccounted for on a species level. 
		Although methods to estimate detection probability have been developed and implemented for many North American birds, estimates are most accurate for common, easy to detect species. 
		There is therefore a need for methodology that provides detection estimates for species that are rare or undersampled, which generally lack sufficient data for accurate estimates. 
		Here, based on results of a previous study, we take advantage of similarities in detection probabilities among phylogenetically-related species and species with similar traits to estimate detection probabilities in seven species of rare North American birds, using hierarchical Bayesian multi-species removal and distance models. 
		We showed that a multi-species model improved precision of detectability estimates for species with low sample size, and even with species up to 1000 observations.
		However, when no data were available for a species, the multi-species models presented here tended to be overly conservative in predicting detectability. 
		We recommend using these models to 1) provide more precise estimates of detectability for species where there are sufficient data, 2) generate estimates of detectability for species for which we have sparse (but non-zero) data, and 3) improve upon previous estimates of detection probabilities for North American birds by combining this model with survey-level covariates.

@@

\section{Keywords}
\par hierarchical modelling, distance sampling, removal sampling, QPAD, detectability, effective detection radius, cue rate, phylogenetics, Bayesian

@@

\par Evidence suggests detectability in birds is at least partially driven by variation in species traits.
\citet{johnston_species_2014} showed that detection distance---and consequently perceptibility---varied with body mass and habitat association in a selection of European landbirds.
In this case, larger body size was associated with longer detection distances, likely because  larger body sizes are easier to detect visually and typically associated with louder and lower-pitched songs \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
Moreover, vegetation structure influences the attenuation of sound, whereby sound diminishes quicker in a closed-forested habitat than more open habitats, thus leading to variation in detectability among species associated with different habitats \citep{waide_tropical_1988, yip_sound_2017}.
\citet{solymos_phylogeny_2018} considered the effects of species traits and phylogeny on both detection distance and cue rate, the latter of which influences the availability of birds for detection.
They found that cue rate is most related to phylogeny, and, to a lesser extent, migratory strategy (i.e., whether a bird was a long-distance migrant or a resident species).
They also found that detection distance was related to species’ traits such as body mass, song pitch, habitat association, and that after accounting for these traits, phylogeny explained very little.

\par \citet{solymos_phylogeny_2018} recommended using these phylogenetic and trait relationships to predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) more broadly.
 By leveraging information provided by species with many detections in available datasets, it may be possible to inform detectability estimates for species with fewer (including potentially zero) detections \citep{sollmann_hierarchical_2016, pacifici_guidelines_2014, zipkin_impacts_2009, white_conservation_2013}.
This could be done in two ways: by making use of shrinkage factors and partial pooling, as done in Bayesian hierarchical models \citep{gelman_what_2021, gelman_bayesian_2006}, and by accounting for phylogenetic relationships and the relative relatedness of different species, as done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}.

\par The goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of sample size requirements for estimating detectability, and to derive estimates of detectability for bird species without any data.
To achieve this goal, we aimed to satisfy the following objectives:
\begin{enumerate}
	\item Compare the mean and precision of estimates of cue rate and EDR produced from a multi-species, Bayesian hierarchical QPAD removal model and distance model, respectively, to estimates produced from single-species models that do not share information across species.
	For species with sufficient sample sizes, we predicted the mean and precision of cue rate and EDR estimates will be similar from multi- and single-species models \citep{edwards_point_2023}.
	However, for species with relatively small sample sizes, we predict estimates for both cue rate and EDR will be more precise from the multi-species Bayesian model.
	\item Assess the capacity for a multi-species model to produce estimates of cue rate or EDR for species that are data sparse.
	We predicted that a multi-species model will be able to produce better fitting estimates of cue rate and EDR than a single species model, particularly for data sparse species.
	\item Assess the capacity for a multi-species model to make out-of-sample predictions of cue rate or EDR for species which have no data.
	We predicted that a multi-species model will be able to produce estimates of cue rate or EDR that fall within a reasonable range of estimates from related species.
\end{enumerate}

We then provide a case study of predicting detectability components for seven species for which we had little to no data: Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Spotted Owl (\textit{Strix occidentalis}), Bicknell’s Thrush (\textit{Catharus bicknelli}), LeConte’s Thrasher (\textit{Toxostoma lecontei}), Harris’s Sparrow (\textit{Zonotrichia querula}), Tricolored Blackbird (\textit{Agelaius tricolor}), and Kirtland’s Warbler (\textit{Setophaga kirtlandii}). 	
All seven of these bird species are listed on Partners in Flight’s watchlist, indicating their conservation importance \citep{will_handbook_2020}.

@@

\subsection{Data Collection}
\subsubsection{Point Count Data}
\par We used the same standardized database of point count data that was assembled through the NA-POPS project \citep{edwards_point_2023}. 
This consists of $712,138$ point counts conducted across $292$ projects in Canada and the United States, $422,514$ of which used removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and $522,820$ of which used distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}.
 All point counts were recorded in the database using a standard method developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.

@@

\subsubsection{Species Traits}
\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts for all species considered in this study \citep{billerman_birds_2022}. 
Species were classified as "closed habitat" if their primary habitat was forests, and "open habitat" if their primary habitat was grasslands or shrublands.
Species with specific areas of their range that are only for breeding (i.e., not year-round throughout their range) were classified as "migratory".
We note that for the purposes of this study, altitudinal and latitudinal migrants were classified as migratory, but irruptive species such as crossbills were not.
Body mass for each bird species was obtained from the Elton traits dataset if available \citep{wilman_eltontraits_2014}, and otherwise through the Birds of the World account \citep{billerman_birds_2022} if not.

@@

\par We will first begin by describing the point count data that are to be used in the models.
For each point count or sampling event $i \in [1, I]$, where $I$ is the number of sampling events in the study, we have recorded observed abundances $Y$ for species $s \in [1,S]$, where $S$ is the number of species in the study.
If a point count used removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002}, then these recorded observed abundances will be further subdivided into a time band $j \in [1,J_i]$, where $J_i$ is the maximum number of time bins for sampling event $i$.
Further, if a point count used distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}, then these recorded observed abundances will also be further subdivided into a distance bin $k \in [1,K_i]$, where $K_i$ is the maximum number of distance bins for sampling event $i$.
Thus, we let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occurring in time band $j$ and/or distance band $k$.

@@

$${Y}_{sij.} \sim \mathrm{multinomial}\left({Y}_{si..}, \mathbf{\Pi}_{si}^{\left\{\phi\right\}}\right).$$

\par For sample $i$, $\mathbf{\Pi}_{si}^{\left\{\phi\right\}}$ is the corresponding probability vector that determines how the total individuals of species $s$ observed at sample $i$ (i.e., ${Y}_{si..}$) are to be distributed across the $J$ time bands; in essence, it is the mixing parameter.
Let $t_{ij}$ be the maximum time for time band $j$ during sampling event $i$, and let $\phi_s$ be the unknown cue rate for species $s$.
We then have the following calculation of each component $\pi_{sij}^{\left\{\phi\right\}}$ of the mixing parameter $\mathbf{\Pi}_{si}^{\left\{\phi\right\}}$:

\begin{equation*}
	\pi_{sij}^{\left\{\phi\right\}} = 
	\begin{cases}
		\dfrac{\exp\left\{ -t_{i,j-1}\phi_{s} \right\} - \exp\left\{ -t_{ij}\phi_{s} \right\}}{1 - \exp\left\{ -t_{iJ}\phi_{s} \right\}} & \text{for } j > 1 \\
		1 - \sum_{n = 2}^{J} \pi_{sin}^{\left\{\phi\right\}} & \text{for } j = 1
	\end{cases}
\end{equation*}

\par We are most interested in estimating $\phi_s$, the cue rate for species $s$. 
To allow for the partial pooling of cue rate estimates for each species to be based on a phylogenetic relationship, we can draw estimates of cue rate from a multivariate normal distribution. 
Let $\mathbf{\Phi}$ be the vector of cue rates (of length $S$), where each element of $\mathbf{\Phi}$ corresponds to a species' cue rate $\phi_s$.
We then have:

$$\log \mathbf{\Phi} \sim N_{[S]}\left( \mathbf{\mu}_{[S]}, \mathbf{\Sigma}_{[S \times S]} \right),$$
where $N_{[S]}(\cdot,\cdot)$ denotes a multivariate normal distribution of length $S$.
The covariance function $\mathbf{\Sigma}$ (denoted with subscript $[S \times S]$ to show dimensionality) determines the strength of the partial pooling, and can be any function that contains a measure of distance. 
In this case, we will create a function based on phylogenetic distance $\mathbf{C}_{[S \times S]}$, i.e.,

$$\mathbf{\Sigma}_{[S \times S]} = \mathbf{C}_{[S \times S]}\lambda\sigma.$$
Here, $\lambda$ is Pagel's Lambda \citep{pagel_inferring_1999}, which determines the strength of the phylogenetic relationship. 
We set $\lambda = 0.76$, consistent with previous results \cite{solymos_phylogeny_2018}. 
We could have chosen to draw estimates from a distribution of possible $\lambda$ as in \citet{solymos_phylogeny_2018}, but we chose not to for simplicity. 
We note that the suite of species modelled in \citet{solymos_evaluating_2018} was primarily Boreal birds, and so there could be slight differences in the phylogenetic strength of a larger suite of species such as what we model here.
We also estimated the extra variance term and used an exponential prior $\sigma \sim exp(5)$.
Finally, the vector of mean cue rate $\mathbf{\mu}_{[S]}$ is a function of the species migratory strategy $\alpha_M$, where the migratory strategy $M$ could either be resident or migrant. 
We set priors on both migratory strategies to be $\alpha_M \sim N(-1, 0.01)$, because on the log scale we would expect cue rate to be negative.

@@

\par For distance modelling, we are only interested in the distance to the cueing bird, and not the time bin that the bird was recorded in, and so we sum counts from each sampling event $i$ over all time bands. 
Thus, we consider counts ${Y}_{si.k} = \sum_{j=1}^{J}{Y}_{sijk}$. 
From \citet{solymos_calibrating_2013}, we model ${Y}_{si.k}$ as
$${Y}_{si.k} \sim multinomial\left({Y}_{si..}, \mathbf{\Pi}_{si}^{\left\{\tau\right\}}\right).$$

\par For sample $i$, $\mathbf{\Pi}_{si}^{\left\{\tau\right\}}$ is the corresponding probability vector that determines how the total number of individuals of species $s$ observed at sample $i$ (i.e., ${Y}_{si..}$) are to be distributed across the $K$ distance bins; in essence, it is the mixing parameter. 
Let $r_{ik}$ be the maximum distance for distance bin $k$ during sampling event $i$, and let $\tau_s$ be the unknown effective detection radius for species $s$. 
We then have the following calculation of each component $\pi_{sik}^{\left\{\tau\right\}}$ of the mixing parameter $\mathbf{\Pi}_{si}^{\left\{\tau\right\}}$:

\begin{equation*}
	\pi_{sik}^{\left\{\tau\right\}} = 
	\begin{cases}
		\dfrac{f(r_{i,k}, \tau_s) - f(r_{i,k-1}, \tau_s)}{f(r_{i,K}, \tau_s)} & \text{for } k > 1 \\
		1 - \sum_{n = 2}^{K} \pi_{sin}^{\left\{\tau\right\}} & \text{for } k = 1
	\end{cases}
\end{equation*}

where 
$$f(r,\tau) =  1 - \exp\left\{ -\dfrac{r^2}{\exp\left\{\tau^2\right\}} \right\}.$$

@@

\par In this model, mean effective detection radius for species $s$ (i.e., $\mu_s$) is a function of the following traits as described in \cite{solymos_phylogeny_2018}: an overall intercept term $\alpha_0$; migratory strategy $\alpha_M$, where $M$ can be migratory or resident; habitat preference $\alpha_H$, where $H$ can be open or closed habitat; log body size $\beta_b$; and log song pitch $\beta_p$. 
That is, we have
$$ \mu_s = \alpha_0 + \alpha_{M_s} + \alpha_{H_s} + \beta_b \times \log BodySize_s + \beta_p \times \log SongPitch_s.$$

We then set the following priors:
$$\alpha_0 \sim N(0.05, 0.1)$$
$$ \alpha_M, \alpha_H \sim N(0, 0.05)$$
$$ \beta_b \sim N(0.01, 0.005)$$
$$ \beta_p \sim N(-0.01, 0.005) $$
$$\sigma \sim exp(5)$$

@@

\subsection{Model Assessment}
\subsubsection{Objective 1: Single-Species vs. Multi-Species Estimates}
\par To test our prediction that there would be little change between estimates of species with sufficient sizes, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data.
We did not include Woodhouse's Scrub-jay (\textit{Aphelocoma woodhouseii}), Pacific Wren (\textit{Troglodytes pacificus}), or Sagebrush Sparrow (\textit{Artemisiospiza nevadensis}) in the analysis, because these species were not available in the phylogenies from \citet{jetz_global_2012}.
Rufous Hummingbird (\textit{Selasphorus rufus}) was dropped from the distance modelling due to issues with divergent transitions (\citet{betancourt_diagnosing_2016, leimkuhler_simulating_2005}; see Discussion).
Tricolored Blackbird was only included in the distance model because there were only sufficient distance sampling data.
This resulted in a total of 316 species for removal modelling, and 315 species for distance modelling.
We also ran the single species removal and distance models for the same suite of birds.
We then modelled the multi species estimates of cue rate and EDR as functions of the single species estimates of cue rate and EDR, and obtained estimates of slope and y-intercept for this relationship.
If our prediction is true, we would expect to see a slope of 1 and a y-intercept of 0, indicating a 1-to-1 relationship.

@@

\par To test our prediction about improved accuracy, we ran a k-fold cross-validation with 5 folds, stratified by species.
That is, for each of the five cross-validation runs, we held out 20\% of the data for each species for both the multi-species and single-species models and ran the models with the remaining 80\% of the data. 
We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR, for each cross-validation run, and used these parameter estimates to calculate the estimated log pointwise posterior density (LPPD) of each of the held out data points, which tells us the likelihood of the data point under the estimated parameters.
We then calculated the difference in LPPD for removal and distance models, by subtracting the pointwise LPPD of the single-species model from the LPPD of the multi-species model (i.e., multi-species minus single-species).
Finally, we took the mean and standard error of the difference in LPPD for each species to create a species-specific mean and standard error of difference in LPPD.
With this, positive intervals (i.e., where mean $\pm$ standard error is all positive) would indicate that the multi-species model was preferred, and negative intervals would indicate that the single-species model was preferred, because there would be an overall higher average likelihood of the data points given the parameters in the multi-species model than the single-species model with a positive interval.
If our prediction is true, then we would expect to see predominantly positive values of the species-specific mean LPPD for species with lower sample sizes.

\subsubsection{Objective 3: Assessment of Out-of-sample Predictions}
\par To test our prediction that a multi-species model would be able to predict cue rates or EDRs for species with no data, we ran a 10-fold leave-species-out cross validation.
That is, for each of the ten cross-validation runs, we held out all data for 10\% of the species for both the multi-species and single-species models.

\par To determine which species would be left out in each fold of the removal model, we first split the phylogenetic tree into 10 parts using the ``cutree" function in base R, which preserves relatedness in the 10 parts of the split tree.
Then, for each of the parts of the tree, we randomly assigned a number between 1 and 10 (drawn from a uniform distribution) separately for all migratory and then all resident species.
This ensured that for each fold, there would be a balance of related and unrelated species left out, as well as a relatively equal balance of migratory and resident species left out.

\par To determine which species would be left out in each fold of the distance model, we randomly assigned a number between 1 and 10 (drawn from a uniform distribution) to each species within a migratory status $\times$ habitat preference grouping combination, which ensured that there would be a good balance of species left out for each group of species.

For each cross-validation fold, we drew values from the posterior distribution of the held-out species, and used mean and standard deviation of those draws as the predicted cue rate or EDR.
We then modelled the predictions of cue rate and EDR obtained with no data as a function of cue rate and EDR obtained with the full data, and obtained estimates of slope and y-intercept for this relationship.
If our prediction is true, then we would expect to have a slope of 1 and a y-intercept of 0, indicating a 1-to-1 relationship.

\subsection{Case Study: Predicting Detection Probabilities for Rare and Undersampled Species}
\par We ran the multi-species removal and distance models again, this time including the minimal data available for seven species of undersampled landbirds, and generated estimates of cue rate and EDR for these seven species.
For Tricolored Blackbird, since we had sufficient data points for distance modelling for Objectives 1 - 3, we only generated estimates of cue rate here.
For two of the species, Kirtland's Warbler and Bicknell's Thrush, we used the predicted cue rate and EDR to calculate availability and perceptibility, and hence detectability, for a given survey type, and compared these metrics to previous estimates of detectability obtained from other studies.
As a reminder, following derivations in \citet{solymos_calibrating_2013}, detectability is the product of availability and perceptibility.
Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as

@@

\par Out-of-sample predictions of EDR also tended to be underestimated compared to estimates obtained with full data (y-intercept = $4.46$, slope = $0.72$; Figure \ref{fig:species_cv}B).
Of particular note was the fact that the model did not generate any out-of-sample predictions greater than 115 m, nor less than 69 m.
In fact, the model was particularly conservative in generating estimates for each of the migrant $\times$ habitat groups, in that there was very little variation within each of the groups; any variation that did occur would have been due to differences in song pitch and body size (Figure \ref{fig:distance_cv_traits}).
